\begin{acknowledgements}

    Like any journey worth taking, a PhD is never traveled alone.
    %
    This dissertation represents not only years of research, but also the culmination of support, mentorship, and kindness from so many people who have shaped my path as a physicist.
    %
    My adventure in physics began with an incidental chat with Vipul Mehra over lunch in my first year of high school.
    %
    Thank you, Vipul, for pushing me harder than I thought I could go, and believing in me when I didn't believe in myself.
    %
    At Yale, I was fortunate to learn from some of the greatest minds in the field.
    %
    I want to thank Witold Skiba, for teaching me both how to do physics and how to do academia.
    %
    I still find myself wishing I was in your classroom furiously keeping up with you before you pause and proclaim ``This is not a fit.''
    %
    I also want to thank Laura Newburgh, David Poland, Tom Appelquist and Vincent Moncrief for guiding me patiently through my first research experiences.
    
    Since arriving at Berkeley and LBNL, I have been surrounded by brilliant and generous colleagues.
    %
    This PhD would not have been possible without the love and support of the Nachman ML Group.
    %
    To the postdocs, Aishik Ghosh, Dennis Noll, Vinicius Mikuni, Sascha Diefenbachar, Mariel Pettee, Fernando Torales Acosta and Tanvi Wamorkar, you made the hardest days easy.
    %
    To my co--grad students, Radha Mastandrea and Jordan Dudley, you have always been a source of inspiration.
    %
    Over the course of this PhD, I have had the privilege of working with some of the most brilliant, generous and thoughtful scientists in the field, and I apologise to all of you who have been part of this journey who I have not thanked here explicity.
    %
    To all of my collaborators---Jesse Thaler, Fernando Torales Acosta, Vinicius Mikuni, Jingjing Pan, Richard Zhu, Mikael Kuusela, Larry Wasserman, and Owen Long---thank you for all the long conversations, your incredibly patience with my strangly timed messages, and your willingness to walk me through your thought processes.
    %
    I am grateful to my dissertation and qualifying exam committee members, Joshua Bloom, Saul Perlmutter and Hitoshi Murayama, for their guidance in charting this journey and sharing their wisdom and unique perspectives that were instrumental in making it a success. In particular, I would like to thank Uros Seljak for being an invaluable co--advisor.

    Words feel inadequate to express my gratitude to Benjamin Nachman.
    %
    As his first graduate student, I've had the privilege of not just an advisor, but a true mentor and friend.
    %
    Ben has believed in me when I struggled to believe in myself, celebrated my successes as if they were his own, and somehow always known exactly when I needed encouragement or a gentle push forward.
    %
    His unwavering support, boundless enthusiasm for physics and genuine care have shaped not just this dissertation but who I am as a scientist and person.
    %
    I have often told new PhD students that should find themselves and advisor who makes them feel like they are the best version of themselves.
    %
    Ben has been all that and more;
    %
    I am profoundly lucky to have had him as my advisor.
    
    My sincere appreciation also goes to the administrative and technical staff of the Physics Department at the University of California, Berkeley, and Lawrence Berkeley National Laboratory for their invaluable assistance.
    %
    Additionally, I acknowledge the support from funding agencies and computational resources provided by the university and national lab.
    %
    Large language models were occasionally consulted during the preparation of this dissertation for brainstorming structural elements, refining transitions between sections and surveying relevant literature.
    %
    However, the presented ideas, analyses and conclusions are entirely my own.
    %
    I attest that I stand by all the facts and references presented in this dissertation.
    %
    I am expecially grateful to the National Energy Research Scientific Computing Center (NERSC) for providing the computational resources that made this work possible.
    %
    This adventure would not have been possible without the Berkeley physics community, that made a home for me far away from home.
    
    Finally, to my parents, Bijal and Nilesh, and my brother, Dweej, there aren't enough words in this entire disseration to express how grateful I am.
    %
    And to the friends who have loved me like family, Tyler, Charlie, Kevin, Benjamin, Lenny, Simon, Sam, Narain, Fran, Levi, David, Jeremy, Anna, Gabe and so many more, here's to many more smiles and laughs to come.

    This disseration is not the end of a road, it merely marks the beginning of the next adventure.
    %
    I am grateful to everyone who was part of this journey for helping me reach this milestone, and to everyone who will be part of the next.
    
\end{acknowledgements}