\begin{longtable}{@{}m{0.18\linewidth}m{0.27\linewidth}m{0.49\linewidth}@{}}
    \caption[Symmetry breaking induced by detector effects in HEP experiments.]{Ideal SM symmetries and detector induced symmetry breaking in HEP experiments.
    %
    While fundamental interactions may respect certain symmetries exactly,  detector geometries, material distributions and reconstruction algorithms violate these symmetries at measurement level.
    %
    The table illustrates how common detector limitations transform SM symmetries.
    %
    These effects must be carefully modelled in simulations and corrected through calibration to extract the underlying physics.
    }
    \label{tab:detectorSym} \\
    \toprule
    \textbf{Symmetry} &
    \textbf{Ideal outcome} &
    \textbf{Detector effect} \\
    \midrule
    \endfirsthead
    
    \multicolumn{3}{c}{\textit{Table \ref{tab:detectorSym} continued from previous page}} \\
    \toprule
    \textbf{Symmetry} &
    \textbf{Ideal outcome} &
    \textbf{Detector effect} \\
    \midrule
    \endhead
    
    \midrule
    \multicolumn{3}{r}{\textit{Continued on next page}} \\
    \endfoot
    
    \bottomrule
    \endlastfoot
    
    \textbf{Azimuthal rotation} $\phi \to \phi + \alpha$ &
    Rotational invariance: uniform distribution in $\phi \in [0, 2\pi]$. No preferred transverse direction &
    Discrete $n$-fold symmetry: detector segmentation creates $\phi-$dependent acceptance.
    %
    Dead material between modules introduces periodic inefficiencies.
    %
    Triggers based on detector regions further break symmetry \\
    
    \textbf{Lorentz boost} $\beta_z \to \beta_z + \Delta\beta$ &
    Physics invariant under longitudinal boosts; cross sections expressible in terms of Lorentz scalars. &
    Lab frame dependence: fixed detector geometry defines preferred frame. Forward boosts push particles beyond acceptance. Resolution degrades for highly boosted objects. Trigger thresholds defined in lab frame are not invariant \\
    
    \(\mqty*{\textbf{Parity}\\ \vb x \to -\vb x}\) &
    For P conserving processes: equal rates and distributions for original and parity transformed events &
    Geometric asymmetry: detectors rarely possess reflection symmetry about $z=0$. Forward and backward regions have different instrumentation, acceptance, and resolution. Magnetic field direction picks out handedness \\
    
    \(\mqty*{\textbf{Charge}\\\textbf{conjugation} \\q \to -\bar q}\) &
    C symmetric interactions produce particles and antiparticles with identical rates and kinematic distributions &
    Charge dependent efficiency: opposite charges bend oppositely in magnetic field, sampling different detector regions. Material interactions differ (e.g., $K^+$ vs $K^-$ nuclear cross sections). Trigger and particle ID algorithms may have charge bias \\
    
    \textbf{Permutation Symmetry} $(i \leftrightarrow j)$ &
    For identical particles: joint distribution exhibits exchange symmetry &
    Position dependent response: particles in different regions experience different resolutions, efficiencies, and systematics. Ordering by $p_T$ obscures underlying symmetry. Combinatorial background differs for same region vs different region pairs \\
\end{longtable}